---
title: "Bayesian Imputation for Censored Recurrent Events"
author: "Patrick M Schnell and Xin Liu"
format: html

vignette: >
  %\VignetteIndexEntry{Data formatting and model fitting example}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(bicre)
```

## Data format

```{r}
data("data_sim")

data_cov <- data_sim$data_cov
data_events <- data_sim$data_events
```

The `data_sim` dataset is a 30-individual simulated dataset generated following the simulation study configuration in Liu & Schnell "Bayesian Data Augmentation for Recurrent Events Under Intermittent Assessment in Overlapping Intervals with Applications to EMR Data" using the "Baseline" scenario settings.
It contains two data frames:

  1. `data_events` which contains the censored recurrent event observations;
  2. `data_cov` which contains time-dependent covariate data.
  
The names of these datasets are flexible, as long as they meet the formatting requirements described below.

### Event data

```{r}
data_events |>
  head()
```
  
The columns of `data_events` are:

  - `id`: A subject-level identifier which links data rows from the same subject within `data_events` and between `data_events` and `data_cov`;
  - `t_start` and `t_end`: The start and end points of the interval represented by each row, open on the left, closed on the right.
  - `e_min` and `e_max`: The minimum and maximum number of events that could have occurred within the interval (i.e., the inclusive endpoints of the censoring interval).
  
That is, each row records that individual `id` experienced a minimum of `e_min` and maximum of `e_max` events in the interval (`t_start`, `t_end`].

The names of these variables are not currently flexible---`bicre` expects exactly these column names with the described meanings.
Additional columns are allowed but ignored.

### Covariate data

```{r}
data_cov |>
  head()
```

Key columns of `data_cov`, which should be included in any dataset properly formatted for `bicre`, are:

  - `id`: A subject-level identifier which links data rows from the same subject within `data_cov` and between `data_cov` and `data_events`;
  - `t_start` and `t_end`: The start and end points of the interval represented by each row.

The other variables (here `trt`, `age`, and `sex`) are covariates, which are assumed to be constant within each interval.
There are no special naming requirements for these other variables.

It is important that the intervals for each `id` partition the follow-up time for that individual, though that follow-up time need not be uninterrupted.
Specifically, any time within an interval (`t_start`, `t_end`] given by a row of `data_events` must be included in exactly one interval (`t_start`, `t_end`] given by a row of `data_cov`.

## Model fitting

The primary function for model fitting is `bicre()`, which provides a formula interface but requires passing two datasets meeting the formatting requirements described above.
The left-hand side of the formula is left empty, while the right-hand side should include the (possibly time-varying) covariates for the model.
A model frame and design matrix is constructed from the formula via `stats::model.frame` and `stats::model.matrix`.

By default, the function saves its progress periodically (by default, every 1000 iterations), though we suppress that behavior here by setting `run_and_save = FALSE`.
Saving progress allows the sampler to be stopped and resumed later by `bicre_continue()`.
Both `bicre()` and `bicre_continue()` have manual entries accessible via, e.g., `?bicre`, which include instructions for specifying priors.

Below, we fit a model with main effects of `sex`, `age`, and `trt`, requesting 100 burn-in iterations and 1000 iterations to retain.
All 1100 iterations are returned, but MCMC proposal distributions are updated only during burn-in.

```{r fit-model}
#| cache: false

fit <- bicre( ~ sex + age + trt,
              data_covariates = data_cov,
              data_events = data_events,
              n_burn_in = 10, n_keep = 100, seed = 0,
              run_and_save = FALSE)
```

The resulting object includes a posterior trace matrix and proposal acceptance rates after burn-in.

```{r}
names(fit)
dim(fit$trace)
colnames(fit$trace)
```

```{r plot}
plot(fit$trace[, "log(b)"])
```

## Data pre-processing

The data as stored in `data_cov` and `data_events` require processing before use in the MCMC sampler.
The `bicre()` function handles this processing automatically if used with the formula interface shown above.
However, for large or complex datasets this processing can require substantial computation time, and it may be advisable to perform these steps before invoking `bicre()`, especially if fitting multiple models or debugging.

Pre-processing proceeds in two major steps:

  1. `co_events()` reorganizes the `data_cov` and `data_events` data frames into a list in which each element corresponds to an individual `id`, and contains `covariates` and `events` data frames specific to that individual;
  2. `build_imputation_units()` partitions each individual's follow-up time into units for which event histories may be imputed independently of all other units.
  
```{r}
ce <- co_events(data_covariates = data_cov,
                data_events = data_events,
                id = id,
                t_start = t_start,
                t_end = t_end,
                e_min = e_min,
                e_max = e_max)

ce[[1]]
```

The `id` through `e_max` arguments indicate which columns in `data_covariates` and `data_events` should be treated as the special columns described in previous sections.

The data frames from `co_events()` may then be converted to model frames by passing the result to `co_events_frame()` with the desired formula.

```{r}
cef <- co_events_frame(ce, formula = ~ sex + age + trt)

cef[[1]]
```

Unless more control is desired, the result of `co_events_frame()` may be passed directly to `build_imputation_units()` without any other arguments.

```{r}
iu <- build_imputation_units(cef)
```

Then, the computed imputation units can replace the data arguments in `bicre()`.

```{r}
#| cache: false

fit <- bicre(iu = iu,
             n_burn_in = 10, n_keep = 100, seed = 0,
             run_and_save = FALSE)
```

```{r}
plot(fit$trace[, "log(b)"])
```
